A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-11 main.tex
Format.
[WIP] revising and adding comments

\newcommand{\G}{\Gamma}
\newcommand{\D}{\Delta}
\newcommand{\V}{\vdash_\Sigma}
\newcommand{\VT}{\vdash\hspace{-.50em}\raisebox{0.28em}{\tiny{$\TB$}}}
\newcommand{\iskind}{\text{\ kind}}
\newcommand{\TW}{\triangleright}
\newcommand{\TWL}{\triangleleft}
\newcommand{\F}{\forall}
\newcommand{\TB}{\blacktriangleright}
\newcommand{\TBL}{\blacktriangleleft}
\newcommand{\E}{\equiv}
\newcommand{\FV}{\text{FV}}
\newcommand{\FTV}{\text{FSV}}

@@

The organization of this paper is the following.
Section 2 gives an informal overview of \LMD.
Section 3 defines the syntax, type system, full reduction, evaluation contexts of \LMD.
Section 4 shows the properties of \LMD including Unique Decomposition.
Section 5 discusses future works.
Finally, Section 6 discusses related works.

@@

In this section, we present \LMD in detail. 
We will define syntax, full reduction, type system including type equivalence rules, values, and evaluation context, which take stages into account.
In the next section, we will discuss on properties of \LMD.

@@

\LMD is defined by the following grammar.

\begin{align*}
    \textrm{Type variables}  &   &                          & X,Y,Z                                                                                                      \\
    \textrm{Variables}       &   &                          & x,y,z                                                                                                      \\
    \textrm{Stage variables} &   &                          & \alpha,\beta,\gamma                                                                                        \\
    \textrm{Stage}           &   &                          & A,B,C                                                                                                      \\
    \textrm{Kinds}           &   & K,J,I,H,G                & ::= * \mid \Pi x:\tau.K                                                                                    \\
    \textrm{Types}           &   & \tau,\sigma,\rho,\pi,\xi & ::= X \mid \Pi x:\tau.\tau \mid \tau\ M \mid \TW_{\alpha} \tau \mid \F\alpha.\tau                          \\
    \textrm{Constants}       &   &                          & c                                                                                                          \\
    \textrm{Terms}           &   & M,N,L,O,P                & ::= c \mid x \mid \lambda x:\tau.M\ \mid M\ M \mid \TB_\alpha M                                            \\
                             &   &                          & \ \ \ \ \mid \TBL_\alpha M \mid \Lambda\alpha.M \mid M\ \epsilon \mid \%_\alpha M                          \\  
    \textrm{Signature}       &   & \Sigma                   & ::= \emptyset \mid X::K \mid c:\tau                                                                        \\
    \textrm{Contexts}        &   & \Gamma                   & ::= \emptyset \mid  \Gamma,x:\tau @A\ (x\not\in\textrm{Domain}(\G)\ \text{and}\ x\not\in\textrm{FV}(\tau)) \\
\end{align*}

@@

The metavariables $X, Y$, and $Z$ range over a countably infinite set of type variables.
$x, y$, and, $z$ range over a countably infinite set of variables.
The metavariables $\alpha, \beta, \text{and} \gamma$ range over the stage variables.
A transition, denoted by $A$ and $B$, is a finite sequence of stage variables.
$\epsilon$ is a symbol for the empty transition.

@@

A kind is either $*$ or $\Pi x:\tau.K$.
Types of terms have kind $*$ and dependent types have $\Pi$-kinds.
For example, $\lambda x:\I.x$ has type $\Pi x:\I.\I$, which has $*$ kind.
\red{この段落は短いので型の段落と結合するか?}

% Types

A type is a type variable which is declared in the signature, a dependent function type,
an application of a term to a dependent function type, a code type, or an $\alpha$-closed type.
A function type in ordinary type systems is omitted 
because it is a special case of a dependent type $\Pi x:\tau$, which has no free occurrence of $x$  in $\tau$.
An application of a term to a dependent function type is, for example, $\text{Vector}\ 10$
if type variable $\text{Vector}$ has the kind $\Pi x:\I.*$.
A code type $\TW_\alpha \tau$ denotes a code fragment of a term of type $\tau$.
An $\alpha$-closed type corresponds to a runnable code fragment.
Type constructors tie strongly in the order of 
an application of a term to a dependent function type $\tau\ M$,
a code type $\TW_{\alpha} \tau$,
a dependent function type $\Pi x:\tau.\tau$, 
an $\alpha$-closed type $\F\alpha.\tau$.
Therefore, $\F\alpha.\TW_{\alpha}.\Pi x:\I.\text{Vector}\ 5$ is interpreted as
$\F\alpha.(\TW_{\alpha}.(\Pi x:\I.(\text{Vector}\ 5)))$.

@@

Terms include a constant $c$ which is declared in signature $\Sigma$.
In addition to ordinary terms in Simply Typed Lambda Calculus, there are five more forms.
$\TB_\alpha M$ represents a code fragment, and $\TB_\alpha M$ represents escape.
$\Lambda\alpha.M$ is a stage variable abstraction.
$M\ \epsilon$ is an application of $\epsilon$ to a stage variable.
$\%_\alpha M$ is a primitive operator for cross-stage persistence.
$\TB_\alpha, \TBL_\alpha, \%_\alpha$ tie stronger than the two forms of applications
and applications are left-associative
and two abstractions extends as far to the right as possible.
For example, $\F\alpha.\lambda x:\I.\TB_\alpha x\ y$ means $\F\alpha.(\lambda x:\I.(\TB_\alpha x)\ y)$.

@@

Signatures are sequences of pairs of a type variable and its kind or a constant and its type.
Because we adopted a \LLF-like system, constants and type variables for base types are given in the signature $\Sigma$.
For example, when we use integers in \LMD, $\Sigma = \I :: *, 1:\I, 2:\I, \cdots$.

% Free variables
As usual, the variable $x$ is bound in $\lambda x:\tau.M$
and the transition variable $\alpha$ is bound in $\Lambda \alpha.M$.
We identify $\alpha$-convertible terms and assume the names of bound variables are pairwise distinct.
We write $\FV(M)$ and $\FTV(M)$ for the set of free variables and the set of free stage variables in $M$, respectively.

@@

\subsection{Reduction}

>>>>

\newcommand{\G}{\Gamma}
\newcommand{\D}{\Delta}
\newcommand{\V}{\vdash_\Sigma}
\newcommand{\VT}{\vdash\hspace{-.50em}\raisebox{0.28em}{\tiny{$\TB$}}}
\newcommand{\iskind}{\text{\ kind}}
\newcommand{\TW}{{\mathop{\triangleright}}}
\newcommand{\TWL}{{\mathop{\triangleleft}}}
\newcommand{\F}{\forall}
\newcommand{\TB}{{\mathop{\blacktriangleright}}}
\newcommand{\TBL}{{\mathop{\blacktriangleleft}}}
\newcommand{\E}{\equiv}
\newcommand{\FV}{\text{FV}}
\newcommand{\FTV}{\text{FSV}}

@@

The organization of this paper is the following.
Section 2 gives an informal overview of \LMD.
Section 3 defines the syntax, type system, full reduction, evaluation contexts of \LMD.
Section 4 shows the properties of \LMD including Unique Decomposition.
Section 5 discusses future works.
Finally, Section 6 discusses related works.
\AI{Use $\backslash$label and $\backslash$ref!}
\AI{``Related work'' and ``future work'' (they are uncountable).}

@@

In this section, we give a formal definition of \LMD, including
the syntax, full reduction, and type system.  In addition to full reduction,
in which any redex at any stage can be reduced, we also give staged reduction,
which models execution as programs.

@@

We assume the denumerable set of type variables, ranged over by
metavariables \(X, Y, Z\), the denumerable set of variables, ranged
over by \(x,y,z\), the denumerable set of constants, ranged over by
\(c\), and the denumerable set of stage variables, ranged over by
\(\alpha, \beta, \gamma\).  The metavariables \(A, B, C\) range over
sequences of stage variables; we write \(\varepsilon\) for the empty
sequence.  \AI{I think $\varepsilon$ is a more standard symbol for
  the nonempty sequence.}  \LMD is defined by the following grammar:

\AI{``Transition'' is still used.}

\begin{align*}
    % \textrm{Type variables}  &   &                          & X,Y,Z                                                                                                      \\
    % \textrm{Variables}       &   &                          & x,y,z                                                                                                      \\
    % \textrm{Stage variables} &   &                          & \alpha,\beta,\gamma                                                                                        \\
    % \textrm{Stage}           &   &                          & A,B,C                                                                                                      \\
    \textrm{Kinds}           &   & K,J,I,H,G                & ::= * \mid \Pi x:\tau.K                                                                                    \\
    \textrm{Types}           &   & \tau,\sigma,\rho,\pi,\xi & ::= X \mid \Pi x:\tau.\tau \mid \tau\ M \mid \TW_{\alpha} \tau \mid \F\alpha.\tau                          \\
%     \textrm{Constants}       &   &                          & c                                                                                                          \\
    \textrm{Terms}           &   & M,N,L,O,P                & ::= c \mid x \mid \lambda x:\tau.M\ \mid M\ M \mid \TB_\alpha M                                            \\
                             &   &                          & \ \ \ \ \mid \TBL_\alpha M \mid \Lambda\alpha.M \mid M\ \epsilon \mid \%_\alpha M                          \\  
    \textrm{Signature}       &   & \Sigma                   & ::= \emptyset \mid \Sigma, X::K \mid \Sigma, c:\tau                                                                        \\
    \textrm{Contexts}        &   & \Gamma                   & ::= \emptyset \mid  \Gamma,x:\tau @A\ (x\not\in\textrm{Domain}(\G)\ \text{and}\ x\not\in\textrm{FV}(\tau)) \\
\end{align*}

\AI{Is $O$ used as a metavariable?}
\AI{Domain and FV are not defined (yet).  I think we should introduce the notion of well-formed type environments by prose below and assume every type environment is well formed.}

@@

@@

A kind, which is used to classify types, is either $*$, the kind of
proper types (types that terms inhabit), or $\Pi x\colon\tau.K$, the kind
of type operators that takes $x$ as an argument of type $\tau$ and returns a type
of kind $K$.
% Types
% of terms have kind $*$ and dependent types have $\Pi$-kinds.  For
% example, $\lambda x:\I.x$ has type $\Pi x:\I.\I$, which has $*$ kind.
% \red{この段落は短いので型の段落と結合するか?}
% Types
A type is a type variable $X$, which is declared in the signature with its kind, a dependent function type $\Pi x:\tau_1.\tau_2$,
an application of a type (operator) of $\Pi$-kind to a term, a code type $\TW_\alpha$, or an $\alpha$-closed type $\F\alpha.\tau$.
We sometimes abbreviate $\Pi x:\tau_1.\tau_2$ to $\tau_1 \rightarrow \tau_2$ if
$x$ is not a free variable in $\tau_2$. \AI{This should be mentioned after free variables are introduced.}
An application of a term to a dependent function type \AI{this sounds like $M\, (\Pi x:\tau_1.\tau_2)$} is, for example, $\text{Vector}\ 10$
if type variable $\text{Vector}$ has the kind $\Pi x:\I.*$.
A code type $\TW_\alpha \tau$ denotes a code fragment of a term of type $\tau$.
An $\alpha$-closed type corresponds to a runnable code fragment.
Type constructors tie strongly \AI{Is ``tie strongly'' standard?} in the order of 
an application of a term to a dependent function type $\tau\ M$,
a code type $\TW_{\alpha} \tau$,
a dependent function type $\Pi x:\tau.\tau$, 
an $\alpha$-closed type $\F\alpha.\tau$.
Therefore, $\F\alpha.\TW_{\alpha}.\Pi x:\I.\text{Vector}\ 5$ is interpreted as
$\F\alpha.(\TW_{\alpha}.(\Pi x:\I.(\text{Vector}\ 5)))$. \AI{I don't understand this rule...}

@@

Terms include ordinary explicitly typed \(\lambda\)-terms, constants,
whose type is declared in signature $\Sigma$, and the following five forms
related to multi-stage programming:
$\TB_\alpha M$ represents a code fragment; $\TB_\alpha M$ represents escape;
$\Lambda\alpha.M$ is a stage variable abstraction;
$M\ \epsilon$ is an application of $\epsilon$ to a stage variable \AI{I think the usage of ``application'' is wrong again}; and 
$\%_\alpha M$ is an operator for cross-stage persistence.
The prefix operators $\TB_\alpha, \TBL_\alpha$, and $\%_\alpha$ tie stronger than the two forms of applications
and applications are left-associative
and two abstractions extends as far to the right as possible.
For example, $\F\alpha.\lambda x:\I.\TB_\alpha x\ y$ means $\F\alpha.(\lambda x:\I.(\TB_\alpha x)\ y)$.

@@

We adopt a \LLF-like system, where constants and type variables for base types \AI{``for base types''?} are given in the signature $\Sigma$, which 
are a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use integers in \LMD, $\Sigma = \I :: *, 1:\I, 2:\I, \cdots$.  \AI{It sounds like we need an infinite signature...}

% Free variables
As usual, the variable $x$ is bound in $\lambda x:\tau.M$
and the transition variable $\alpha$ is bound in $\Lambda \alpha.M$.
\AI{What about other binders such as $\Pi$?}
We identify $\alpha$-convertible terms and assume the names of bound variables are pairwise distinct.
We write $\FV(M)$ and $\FTV(M)$ for the set of free variables and the set of free stage variables in $M$, respectively.

@@

\AI{We should note that constants and type variables are not staged.}
\subsection{Reduction}
